\subsection{Installation}\label{installation}

Simply run \texttt{make ENV=ts7200 install}. This will place the
compiled kernel in \texttt{/u/cs452/tftp/ARM/\$(USER)/k.elf} (where
\texttt{\$(USER)} is whatever your username is on the computer which you
are compiling it). You can then load this kernel from redboot using
\texttt{load -b 0x00218000 -h 10.15.167.5 "ARM/\$(USER)/k.elf"}.

\subsection{Emulator}\label{emulator}

You can run the OS on an emulated arm CPU via
\texttt{make ENV=qemu qemu-run}. In order for this to complete
successfully, you will need:

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  A local install of QEMU
\item
  A local install of an arm-none-eabi toolchain
\end{itemize}

\subsection{Tests}\label{tests}

The provided kernel has some basic tests, which you can run with
\texttt{make ENV=qemu test}. (this depends on the emulator, above).

\subsection{Debugging}\label{debugging}

If something goes wrong and you want to investigate deeper, you can run
\texttt{make ENV=qemu qemu-start} in one terminal (this starts qemu in a
suspended state), and then \texttt{make ENV=qemu qemu-debug} in another.
You can then set breakpoints, step, inspect memory, etc- anything you
can normally do with gdb.

\subsection{Special Instructions}\label{special-instructions}

Before running the program on the TS7200, it is necessary to reset both
the TS7200 and the train controller, to avoid messy state left behind by
previous runs. In particular, this has been known to cause bugs with the
sensor reads, where garbage data left in the UART is picked up, and
misinterpreted.

\subsection{Commands}\label{commands}

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  \texttt{tr \textless{}train number\textgreater{} \textless{}speed\textgreater{}}
  sets the speed of a train
\item
  \texttt{sw \textless{}switch number\textgreater{} (c\textbar{}s)} sets
  a the position of a switch to curved or straight
\item
  \texttt{rv \textless{}train number\textgreater{}} reverses a train
\item
  \texttt{q} exits the program
\end{itemize}

The positions of each switch is shown in the ASCII-art map of the train
track. When a sensor is tripped, it is shown both on the ASCII-art map
for as long as it is tripped, as well as to the right, on the list of
recently-fired sensors.
