\documentclass[titlepage]{article}
\usepackage{listings}
\usepackage{hyperref}

\lstset{basicstyle=\footnotesize\ttfamily,breaklines=true}

\begin{document}
\title{Kernel 3}
\author{Justin McGirr (\#20413625), Peter Raboud (\#20437716)}
\maketitle

\section{Instructions}
\input{"../k1/README.tex"}

\section{Design Decisions}
\subsection{Await Interface}
The implementation of await was something that we debated for quite some time,
because we were given multiple choices in implementation, and because there is
a lot of literature on the different approaches to device driver isolation.
Given that we are writing a microkernel, we want, for various reasons, the
kernel itself to be as small as possible. This allows isolation of failures,
easy prioritization of tasks, and more. However, given that we are, in our toy
kernel, only writing device drivers for a \emph{very} small number of
peripherals, creating all the primitives to allow safe and race-free access
to peripherals from driver tasks seemed like it would likely end up being
bigger and more complex than just implementing a HAL in the kernel. For
example, if the user task is responsible for asking the device to deassert
its interrupt, then you need to make sure that you don't enable interrupts
when switching back to the user task, or else the user task will just get
immediately interrupted again, leading to an infinite loop. However, if you
don't enable interrupts, then you can't preempt the user task, losing a
significant advantage of implementing that portion of the device communication
in userland. Thus, at least for this iteration of the kernel, we have chosen
to have the kernel read the data from the device, deassert the interrupt,
and return the relevant data back to the userland \texttt{await()} call.

For now, the interface to \texttt{await()} is \texttt{int await(int evt);}.
However, this will need to be changed to accommodate passing in additional data
to the kernel, in order to to support UART write events.
We considered masking this data into \texttt{evt}.
Since \texttt{evt} is 32 bits wide, and we only need 3 to represent 5 distinct events,
we could stuff an 8-bit character into the upper bits of the \texttt{int}.
However, if we have control over the signature of \texttt{await()}, it's easier to just
add an extra parameter to contain this data.

\subsection{Await Data Structures}
For keeping track of tasks that are blocked awaiting external events, we used
a array of queues, one queue for each different possible event. This allows
constant-time removal and addition of tasks from waiting queues. The latter
is particularly important, as tasks are currently dequeued immediately in the
interrupt handler, where speed is particularly important, because it can
directly impact the worst-case latency of a response. Had a linear scan been
used instead, this could  yield quite bad latency if many tasks were stacked
up waiting for interrupts, particularly if many were stacked up at the same
time.

We currently dequeue all of the waiting tasks when an event occurs. This could
cause some performance problems if many tasks are queued up at once, because
a single interrupt would have to dequeue all of them. However, this is not
currently a problem, because we only ever have a single task at a time
awaiting an event.

\subsection{Clockserver}
As part of this assignment, we implemented a clock server as a user task, which
allows other user tasks to request the current time, or schedule themselves to run
after some delay.
I assume this will be very useful
when we start having trains going around the track, and we will start having
tasks which care about real-world values like time and distance.

To keep track of tasks which requested future wakeup, we need a prority queue
of tasks, ordered by the time at which they need to be woken up.
We chose to use a min-heap for this purpose.
A min heap is reasonably suitable, since there aren't hard bounds on
the keys which the tasks are sorted by in the priority queue.
This made the priority-queue implementation used for the ready queue inappropriate,
since it relies on the priorities being integers in a very small range.

A min-heap allows $\log(n)$ time insertion and removal of elements, with
constant-time querying of the smallest element.
The latter is the most important property, because we have to check the top
element of this queue against the number of ticks which have passed after each tick.
The $\log(n)$ cost to add and remove elements is probably more expensive than it
sounds, since $n$ is relatively small.
However, since this is not currently a performance bottleneck, we chose not to spend
additional time optimizing for a problem which doesn't currently exist.

An additional reason for the choice of this data structure was that we coincidentally
had one from K1.
We had initially thought we might use it for the priority queue,
but instead chose an alternative data structure for that assignment.

\subsection{Interrupt-Triggered Context Switch}
As part of this assignment, we had to write a context switching
routine to be run when an interrupt was triggered.
We chose to implement this as a context switch into the kernel, where the interface to the kernel just looks
like an SVC call, but with a \texttt{SYSCALL\_IRQ} syscall number, where the kernel then
handles the interrupt just as it would a regular syscall. We considered having
an extra pseudo-task to handle interrupts, with it's own stack, which would
automatically get "scheduled" whenever an interrupt was triggered, but decided
that the additional complexity would not be worth the benefits, and additionally
such a design would waste some space --- the IRQ stack would only be used when
IRQs are triggered, and is unlikely to hold much of anything useful the rest
of the time.

The actual context switching function is different depending on whether the
interrupt originated from an SVC call or a IRQ, although almost all of the
source code is shared through the use of assembly macros. Although we considered
trying to fit both context switches into a single block of code through the use
of conditional execution, it seemed like it would complicate things
considerably, not least of which because you would have to find a register to
store that state in. As the context switch is only slightly more than ten
instructions, we opted just to implement the conditions at the assembler level.

% Things we should probably talk about (not a complete list, just a brain dump
% of some things I think we should include)
% - how we queue tasks awaiting events
% - how we chose our await_event() signature
% - the clockserver implementation (choice of min heap to maintain the list of waiting tasks)
% - how the context switch works in the new world order (and why we chose to
%   have 2 separate context switches)

\section{Questions}

The output produced by running the kernel is reproduced below:

\begin{verbatim}
Boot......IO...
tid: 4, interval: 10, round: 0
tid: 4, interval: 10, round: 1
tid: 5, interval: 23, round: 0
tid: 4, interval: 10, round: 2
tid: 6, interval: 33, round: 0
tid: 4, interval: 10, round: 3
tid: 5, interval: 23, round: 1
tid: 4, interval: 10, round: 4
tid: 4, interval: 10, round: 5
tid: 6, interval: 33, round: 1
tid: 5, interval: 23, round: 2
tid: 4, interval: 10, round: 6
tid: 7, interval: 71, round: 0
tid: 4, interval: 10, round: 7
tid: 4, interval: 10, round: 8
tid: 5, interval: 23, round: 3
tid: 6, interval: 33, round: 2
tid: 4, interval: 10, round: 9
tid: 4, interval: 10, round: 10
tid: 5, interval: 23, round: 4
tid: 4, interval: 10, round: 11
tid: 4, interval: 10, round: 12
tid: 6, interval: 33, round: 3
tid: 5, interval: 23, round: 5
tid: 4, interval: 10, round: 13
tid: 7, interval: 71, round: 1
tid: 4, interval: 10, round: 14
tid: 4, interval: 10, round: 15
tid: 5, interval: 23, round: 6
tid: 6, interval: 33, round: 4
tid: 4, interval: 10, round: 16
tid: 4, interval: 10, round: 17
tid: 5, interval: 23, round: 7
tid: 4, interval: 10, round: 18
tid: 6, interval: 33, round: 5
tid: 4, interval: 10, round: 19
tid: 5, interval: 23, round: 8
tid: 7, interval: 71, round: 2
Exiting kernel...
Task 0 ran for 1981447 us
Task 1 ran for 338 us
Task 2 ran for 310 us
Task 3 ran for 1205 us
Task 4 ran for 3266 us
Task 5 ran for 1379 us
Task 6 ran for 928 us
Task 7 ran for 438 us
Task 8 ran for 528 us
Kernel ran for 146629 us
Ran for 2136468 us total
\end{verbatim}

The first line is output produced as the kernel loads up, to provide feedback in the case that it breaks,
and is not significant here.

The lines between the bootup sequence until \texttt{Exiting kernel...} are
printed by the client tasks, as expected.  Because each message is printed out
following a delay, each task doesn't output until $n$ ticks have passed,
allowing the tasks with the smallest delays to print out several lines before
the tasks with the largest delays get a chance to print any line.

At the moment, the tasks print in a stable / predictable order.
This is because the tasks have different delay periods which aren't multiples
of each other. If the resume and wait process were to be instantaneous, this
would mean that the tasks would wake up at $d, 2d, 3d, \ldots$, and since the
GCD of any two $d$'s of the tasks is bigger than the amount of time the program
runs, these numbers never collide.

In this ideal world, a task is supposed to resume at time $t$, and go back to
sleep immediately. However, in reality, it takes $t + \delta$ time to "call"
\texttt{delay()} again, where $\delta$ includes time to execute the printing,
context switch into the kernel, and context switch into the clock server. This
makes the total time taken $t + \delta + d$, when it would ideally be just
$t + d$.

However, since our kernel is reasonably quick, $\delta$ is never larger than a
single clock tick (10ms). Because two user tasks are never scheduled during
the same tick, as shown above, this is sufficiently close to zero that we
actually output the same lines as we would if our kernel was ideal, and we do so
consistently.

\section{Source Code}
The source code is hosted on git, at \url{git.uwaterloo.ca/pgraboud/cs452-kernel}.
The version we wish to submit is on the \texttt{k3} branch, specifically
the commit:
\input{|"git rev-parse HEAD | ./../verbatim"}
We are submitting the following files:
\input{|"cat ../file-list | ./../verbatim"}
\input{|"./../file-list | ./../verbatim"}

\end{document}
